---
title: "pset2_steiner"
author: "erika steiner"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, results='hide'}
knitr::opts_chunk$set(echo = TRUE)

# load packages
library("haven")
library("tidyverse")
library("magrittr")
library('knitr')
library('kableExtra')
# library("broom")
# library("stringr")
# library("sandwich")
# library("lmtest")
# library("car")
# library("stargazer")

# get rid of scientific notation
options(scipen=999)
```

In this problem set, we continue analyzing the FTP program, but we
reinterpret treatment status. In FTP, as in other programs, there was
confusion among respondents about whether they were subject to a time
limit. For the purposes of this problem set, we will consider
respondents who said that they were subject to a time limited as
treated, and those who said they were not as controls. Our goal will be
to estimate the effect of a perceived time limit on welfare receipt,
employment, and the like. The idea is that you cannot be influenced by a
time limit if you don’t believe it applies to you, whether or not it
applies to you in fact. All data and documentation can be found on
Canvas under Problem Set 1.

(1) *The question about the time limit comes from the four-year survey,
    the results of which are contained in the file ftp_srv.dta. Start by
    merging this file to ftp_ar.dta. How many of the original sample
    members were interviewed for the survey?*

```{r import data}
# import admin data
ftp_ar <- read_dta("ftp_ar.dta")
# import survey data
ftp_srv <- read_dta("ftp_srv.dta")

# this ftp_ar.dta file already has all survey questions imported
# you can confirm this by running the following:
# setdiff(colnames(ftp_srv), colnames(ftp_ar))

ftp_merged <- ftp_ar %>% 
# filter out ftp_ar participants who did not participate in the survey
  # using sampleid per data documentation page 13
  filter(.$sampleid %in% (ftp_srv$sampleid))

# count number of sample members interviewed for survey
nrow(ftp_merged)

# page 8 documentation: Only the 2,160 sample members who entered the study between August 1994 and February 1995 were attempted for this survey. The 1,730 people who completed the survey represent 80 percent of those who were attempted and approximately 60 percent of the report sample.

```

(2) *Tabulate the variable fmi2.*

```{r fmi2}
# fmi2: "Is/was there a time limit?"

# create a table of fmi2 responses
ftp_merged %>% 
  # tabulate
  count(fmi2) %>% 
  # per survey page 48, question I2
  # "1" = "Yes". "2" = "No", "8" = "Don't Know"
  mutate(
    fmi2 = case_when(
      fmi2 == 1 ~ "Yes",
      fmi2 == 2 ~ "No",
      fmi2 == 8 ~ "I don't know",
      # Edit from chatgpt - did not know how to re-format NA counts
      # Prompt: how to handle "NA" format in mutate case when argument r
      # Response: Check for NA explicitly: In case_when(), you need to handle NA values explicitly by using the is.na() function
      is.na(fmi2) ~ "Missing"
    )) %>% 
  # make table
  kable(col.names = c("Response", "Frequency"), caption = "(Is/Was) there a time limit on how long you (are/were) allowed to receive cash assistance?")
```

*How many of the survey respondents have valid (i.e., non-missing)
responses?*

1149

*How many thought they were subject to a time limit?*

666

*How many thought otherwise?*

365

*How many were unsure?*

118

*Finally, what explains why the number of valid responses is less than
the number of people interviewed for the survey?* (Hint: examine
variables fmi1 and fmi1a.)

```{r fmi1a}
# fmi1: currently receiving cash assistance
# fmi1a: have received cash assistance

# fmi1 table
ftp_merged %>% 
  # tabulate
  count(fmi1) %>% 
  # per survey page 48, question I1
  # "1" = "Yes". "2" = "No", "8" = "Don't Know"
  mutate(
    fmi1 = case_when(
      fmi1 == 1 ~ "Yes",
      fmi1 == 2 ~ "No",
      fmi1 == 8 ~ "I don't know",
      is.na(fmi1) ~ "Missing"
    )) %>% 
  # make table
  kable(col.names = c("Response", "Frequency"), caption = "Are you currently receiving cash assistance through AFDC, the WAGES program, or TANF?")

# fmi1a table
ftp_merged %>% 
  # tabulate
  count(fmi1a) %>% 
  # per survey page 48, question I1a
  # "1" = "Yes". "2" = "No", "8" = "Don't Know"
  mutate(
    fmi1a = case_when(
      fmi1a == 1 ~ "Yes",
      fmi1a == 2 ~ "No",
      fmi1a == 8 ~ "I don't know",
      is.na(fmi1a) ~ "Missing"
    )) %>% 
  # make table
  kable(col.names = c("Response", "Frequency"), caption = "Have you ever received cash assistance since (RAD)?")
```

Only 230 people reported actively receiving cash assistance. Of the
\~1500 people who did not report actively receiving cash assistance, 565
people said they hadn't received any cash assistance since randomization
(which could have occurred up to two years prior TK). Thus, many
respondents may not have interacted with cash assistance programs
recently and would not how to respond to this question.

(3) *Define a new treatment dummy that is equal to one for people who
    believed they were subject to a time limit and equal to zero for
    those who did not or were not sure. Call it TLyes. Cross-tabulate
    TLyes and the experimental dummy (“e”).*

```{r tlyes}
ftp_merged %<>%  
  # add new treatment dummy TLyes
  mutate(TLyes = 
    case_when(
      # 1 if believed subject to time limit
      fmi2 == 1 ~ 1,
      # 0 if did not or were not sure if subject to time limit
      fmi2 == 2 ~ 0,
      fmi2 == 8 ~ 0
    )
  )

# cross-tabulate TLyes and e
ftp_merged %>% 
  # tabulate
  count(TLyes, e) %>% 
  # enhance readability of TLyes
  mutate(
    TLyes = case_when(
      TLyes == 1 ~ "Answered Yes TL",
      TLyes == 0 ~ "Answered No TL",
    )) %>% 
  # enhance readability of e
  mutate(
    e = case_when(
      e == 1 ~ "e Treatment",
      e == 0 ~ "e Control"
    )
  ) %>% 
  # pivot into crosstab
  pivot_wider(names_from = e, values_from = n) %>% 
  # make table
  kable(caption = "Cross-tabulating TLyes and e")

```

*Discuss confusion about the time limit.*

Approximately 25% of participants who were subject to time limits
incorrectly believed they were not subject to time limits. Similarly,
31% of participants who were not subject to time limits incorrectly
believed they were subject to time limits.

The inaccuracy of these beliefs is very relevant to this study, which
seeks to understand how imposed time limits change the behavior of
welfare recipients. If participants have inaccurate information about
these time limits, their behaviors may differ from what would be
expected if they had a more accurate understanding. Conclusions drawn
from this study should thus consider how this confusion may confound our
understanding of results.

(4) *Using TLyes as your treatment indicator, estimate by OLS the effect
    of the time limit on the number of months of welfare receipt during
    years 1-2 post_RA. To do this, estimate a regression of your
    dependent variable on TLyes and the set of control variables used by
    MDRC in their publications (the ones with the string “cova:” in
    their labels).*

```{r ols-tlyes}

```

*Do you believe these regressions consistentl estimate the effect of the
time limit? Explain.*

(5) Provide conditions under which the experimental dummy would be a
    valid instrument for the time limit.
(6) Estimate the first-stage regression of TLyes on e. Do you have a
    weak-instrument problem?
(7) Use the experimental dummy as an instrument to estimate the effect
    of the time limit on the number of months of welfare receipt during
    years 1-2 post-RA. How do the welfare receipt results compare to the
    OLS estimates from question (4)? Can you summarize the direction of
    bias associated with the OLS estimates?
(8) Explain why you would expect the exclusion condition to fail for the
    above regression.
(9) Now estimate a model with age-group interactions, along the lines of
    what you did in PS 1. Define the same four age groups, and estimate
    the same model, but interact the age-group dummies with TLyes rather
    than e. Estimate the model by OLS. Omit the control variables.
(10) Now estimate the same model, but use interactions between e and the
     age-group dummies as instruments for the interactions between TLyes
     and the age-group dummies. Comment on how estimating the model with
     age-group interactions helps to satisfy the exclusion restriction.
(11) Discuss your estimated effects of time limits and how they compare
     with the estimates from question (9). Can you diagnose why the two
     sets of coefficients differ?
