---
title: "pset2_steiner"
author: "erika steiner"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE, results='hide'}
knitr::opts_chunk$set(echo = FALSE)

# load packages
library("haven")
library("tidyverse")
library("magrittr")
library('knitr')
library('kableExtra')
library("stringr")
library("stargazer")
library("ivreg")


# get rid of scientific notation
options(scipen=999)
```

(1) *The question about the time limit comes from the four-year survey,
    the results of which are contained in the file ftp_srv.dta. Start by
    merging this file to ftp_ar.dta. How many of the original sample
    members were interviewed for the survey?*

```{r import data}
# import admin data
ftp_ar <- read_dta("ftp_ar.dta")
# import survey data
ftp_srv <- read_dta("ftp_srv.dta")

# this ftp_ar.dta file already has all survey questions imported
# you can confirm this by running the following:
# setdiff(colnames(ftp_srv), colnames(ftp_ar))

ftp_merged <- ftp_ar %>% 
# filter out ftp_ar participants who did not participate in the survey
  # using sampleid per data documentation page 13
  filter(.$sampleid %in% (ftp_srv$sampleid))

# count number of sample members interviewed for survey
nrow(ftp_merged)

# page 8 documentation: Only the 2,160 sample members who entered the study between August 1994 and February 1995 were attempted for this survey. The 1,730 people who completed the survey represent 80 percent of those who were attempted and approximately 60 percent of the report sample.

```

(2) *Tabulate the variable fmi2.*

```{r fmi2}
# fmi2: "Is/was there a time limit?"

# create a table of fmi2 responses
ftp_merged %>% 
  # tabulate
  count(fmi2) %>% 
  # per survey page 48, question I2
  # "1" = "Yes". "2" = "No", "8" = "Don't Know"
  mutate(
    fmi2 = case_when(
      fmi2 == 1 ~ "Yes",
      fmi2 == 2 ~ "No",
      fmi2 == 8 ~ "I don't know",
      # Edit from chatgpt - did not know how to re-format NA counts
      # Prompt: how to handle "NA" format in mutate case when argument r
      # Response: Check for NA explicitly: In case_when(), you need to handle NA values explicitly by using the is.na() function
      is.na(fmi2) ~ "Missing"
    )) %>% 
  # make table
  kable(col.names = c("Response", "Frequency"), caption = "FMI2. (Is/Was) there a time limit on how long you (are/were) allowed to receive cash assistance?") %>% 
  # set style 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

*2a. How many of the survey respondents have valid (i.e., non-missing)
responses?*

1149 (Total 1729 minus 580 NA)

*2b. How many thought they were subject to a time limit?*

666

*2c. How many thought otherwise?*

365

*2d. How many were unsure?*

118

*2e. Finally, what explains why the number of valid responses is less
than the number of people interviewed for the survey?* (Hint: examine
variables fmi1 and fmi1a.)

```{r fmi1a}
# fmi1: currently receiving cash assistance
# fmi1a: have received cash assistance

# fmi1 table
ftp_merged %>% 
  # tabulate
  count(fmi1) %>% 
  # per survey page 48, question I1
  # "1" = "Yes". "2" = "No", "8" = "Don't Know"
  mutate(
    fmi1 = case_when(
      fmi1 == 1 ~ "Yes",
      fmi1 == 2 ~ "No",
      fmi1 == 8 ~ "I don't know",
      is.na(fmi1) ~ "Missing"
    )) %>% 
  # make table
  kable(col.names = c("Response", "Frequency"), caption = "FM1. Are you currently receiving cash assistance through AFDC, the WAGES program, or TANF?") %>% 
  # set style 
  kable_styling(bootstrap_options = c("striped", "hover"))

# fmi1a table
ftp_merged %>% 
  # tabulate
  count(fmi1a) %>% 
  # per survey page 48, question I1a
  # "1" = "Yes". "2" = "No", "8" = "Don't Know"
  mutate(
    fmi1a = case_when(
      fmi1a == 1 ~ "Yes",
      fmi1a == 2 ~ "No",
      fmi1a == 8 ~ "I don't know",
      is.na(fmi1a) ~ "Missing"
    )) %>% 
  # make table
  kable(col.names = c("Response", "Frequency"), caption = "FM1a. Have you ever received cash assistance since (RAD)?") %>% 
  # set style 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Only 230 people reported actively receiving cash assistance. Of the
approximately 1500 people who did not report actively receiving cash
assistance, 565 people said they hadn't received any cash assistance
since randomization (which occurred at least 3.5 years prior). Thus,
many respondents may not have interacted with cash assistance programs
recently and would not how to respond to this question.

(3) *Define a new treatment dummy that is equal to one for people who
    believed they were subject to a time limit and equal to zero for
    those who did not or were not sure. Call it TLyes. Cross-tabulate
    TLyes and the experimental dummy (“e”).*

```{r tlyes}
ftp_merged %<>%  
  # add new treatment dummy TLyes
  mutate(TLyes = 
    case_when(
      # 1 if believed subject to time limit
      fmi2 == 1 ~ 1,
      # 0 if did not or were not sure if subject to time limit
      fmi2 == 2 ~ 0,
      fmi2 == 8 ~ 0
    )
  ) %>% 
  # remove rows where TLyes = NA
  # per Ed discussion board #25
  filter(!is.na(TLyes))

# cross-tabulate TLyes and e
ftp_merged %>% 
  # tabulate
  count(TLyes, e) %>% 
  # enhance readability of TLyes
  mutate(
    TLyes = case_when(
      TLyes == 1 ~ "Answered Yes TL",
      TLyes == 0 ~ "Answered No TL",
    )) %>% 
  # enhance readability of e
  mutate(
    e = case_when(
      e == 1 ~ "e Treatment",
      e == 0 ~ "e Control"
    )
  ) %>% 
  # pivot into crosstab
  pivot_wider(names_from = e, values_from = n) %>% 
  # make table
  kable(caption = "Cross-tabulating TLyes and e") %>% 
  # set style 
  kable_styling(bootstrap_options = c("striped", "hover"))

# cor(ftp_merged$TLyes, ftp_merged$e)

```

*3a. Discuss confusion about the time limit.*

Approximately 25% of participants who were subject to time limits
incorrectly believed they were not subject to time limits. Similarly,
31% of participants who were not subject to time limits incorrectly
believed they were subject to time limits.

The inaccuracy of these beliefs is very relevant to this study, which
seeks to understand how imposed time limits change the behavior of
welfare recipients. If participants have inaccurate information about
these time limits, their behaviors may differ from what would be
expected if they had a more accurate understanding.

We should also consider that a person's understanding of time limits may
be related to some unobservable trait that also affects their welfare
use.

Before drawing conclusions from this study, we should thus consider how
this confusion may confound our understanding of results.

(4) *Using TLyes as your treatment indicator, estimate by OLS the effect
    of the time limit on the number of months of welfare receipt during
    years 1-2 post_RA. To do this, estimate a regression of your
    dependent variable on TLyes and the set of control variables used by
    MDRC in their publications (the ones with the string “cova:” in
    their labels).*

```{r ols-tlyes, results = 'asis'}

# create variable which equals number of months on welfare receipt during years 1-2 post_RA

ftp_merged %<>%
  # variable = number of months rec afdc y1-2 (nmrecyr12)
  mutate(nmrecyr12 = 
           # number of months y1 + number of months y2
           krecc2t5 + krecc6t9) %>% 
  # number of months food stamp receipt
  mutate(nmfs12 = 
           # number of months y1 + number of months y2
           krfsc2t5 + krfsc6t9)

# identify covas columns
# pulled from pset1
covas <- ftp_merged %>% 
  # select
  select( 
    # columns where...
    where(~ {
    # the column's label
    column_label <- attributes(.)$label
    # is not null AND contains 'cova' (returns TRUE)
    !is.null(column_label) && str_detect(column_label, 'cova')
    }))  %>% 
  colnames()

# removing "yrearnsq" because it is collinear with "yrearn" and causes a singular matrix issue
covas %<>% keep(~.x !="yrearnsq")

# write formula for nmrecyr12 regressed on TLyes and control variables (covas)
q4formulaafdc <- as.formula(paste(
  # nmrecyr12 regressed on TLyes and...
  "nmrecyr12 ~ TLyes+", 
  # list covas as string, separated by +
  paste(covas, collapse = " + ")))


q4modelafdc <- lm(q4formulaafdc, data = ftp_merged)

# regress nmrecyr12 on TLyes and set of control variables

q4formulafs <- as.formula(paste(
  # fs regressed on TLyes and...
  "nmfs12 ~ TLyes+", 
  # list covas as string, separated by +
  paste(covas, collapse = " + ")))


q4modelfs <- lm(q4formulafs, data = ftp_merged)

# regress nmrecyr12 on TLyes and set of control variables

stargazer(q4modelafdc, q4modelfs, title = "Regression on TLYes Results", column.labels = c("AFDC receipt", "Food Stamp receipt"), type = "html")

```

*4a. Do you believe these regressions consistently estimate the effect
of the time limit? Explain.*

These regressions do not consistently estimate the effect of the time
limit. "TLyes" is not exogenous, as it is not randomly assigned but
rather based on someone's perception of their time limits. "TLyes" may
thus be correlated to some unobservables, such as understanding complex
welfare program communications, which could also impact the outcome
(number of months on welfare). It is also correlated to the experimental
dummy, which will be shown in future questions. Regression estimates of
the effects of the perception of having a time limit are thus unlikely
to be unbiased or consistent.

(5) *Provide conditions under which the experimental dummy would be a
    valid instrument for the time limit.*

The experimental dummy would be a valid instrument for the time limit if
it satisfies the following 4 assumptions: Stable Unit Treatment Value
Assumption (there are no spillover effects between individuals), the
exclusion restriction (the experimental dummy, e, can only impact the
outcome *through* TLyes), the instrument assumption (random assignment
to time limit impacts someone's perception of the time limit), and the
monotonicity assumption (no one believes they are subject to a time
limit only if they don't receive one).

(6) *Estimate the first-stage regression of TLyes on e.*

```{r first stage, results = 'asis'}
# to calculate the first-stage, regress the treatment dummy on the instrument

first_stage <- lm(TLyes ~ e, data = ftp_merged)

stargazer(first_stage, title = "First-Stage Results", type = "html")

```

*6a. Do you have a weak-instrument problem?*

This regression tells us that being randomly assigned to the time limit
(e) increases the probability of perceiving a time limit (TLyes) by
about 44.5 percentage points. This is significant at the 0.01 level,
suggesting strong evidence that e is correlated with TLyes. The
F-statistic is 291.8, which is much larger than 10 and is statistically
significant. This regression confirms that we do not have a weak
instrument problem.

(7) *Use the experimental dummy as an instrument to estimate the effect
    of the time limit on the number of months of welfare receipt during
    years 1-2 post-RA.*

```{r experimental dummy 2SLS, results = 'asis'}
# estimate the effect of the time limit on the number of months of welfare receipt

# write ivformula for nmrecyr12 regressed on TLyes and control variables (covas), using e as instrument
ivformulaafdc <- as.formula(paste(
  # nmrecyr12 regressed on TLyes and...
  "nmrecyr12 ~ TLyes +", 
  # covariates
  paste(covas, collapse = " + "),
  # given e as instrument and...
  " | e +",
  # covariates
  paste(covas, collapse = " + ")
  ))

ivmodelafdc <- ivreg(ivformulaafdc, data = ftp_merged)

ivformulafs <- as.formula(paste(
  # nmfs12 regressed on TLyes and...
  "nmfs12 ~ TLyes +", 
  # covariates
  paste(covas, collapse = " + "),
  # given e as instrument and...
  " | e +",
  # covariates
  paste(covas, collapse = " + ")
  ))

ivmodelfs <- ivreg(ivformulafs, data = ftp_merged)


stargazer(q4modelafdc, ivmodelafdc, q4modelfs, ivmodelfs, title = "OLS vs IV 2SLS", type = "html")
```

*7a. How do the welfare receipt results compare to the OLS estimates
from question (4)? Can you summarize the direction of bias associated
with the OLS estimates?*

The OLS estimates from question 4 are positive, indicating that people
who believed they were under a time limit would see an increase in
number of months receiving welfare. The IV results are negative,
indicating that a perceived time limit on welfare would reduce the
number of months receiving welfare. The standard error is also much
larger for the IV.

This implies that OLS biases upwards, as the estimate is more positive
than the IV.

(8) *Explain why you would expect the exclusion condition to fail for
    the above regression.*

The exclusion restriction requires that Z (the experimental dummy)
cannot impact the outcome directly, but rather through D (TLyes) alone.
However, this is likely not the case. Even if someone does not perceive
themselves to be under the time limit, the time limit would still limit
the number of months that the participant receives welfare.

As relevant to this problem set, there may also be heterogenous
treatment effects. The IV and OLS models above don't allow for differing
effects among different subgroups. As such, the instrument is not fully
exogenous as it may include additional effects due to subgroup
characteristics.

The experimental dummy is thus endogenous and/or would directly impact
the outcome outside of TLyes, it therefore fails the exclusion
condition.

(9) *Now estimate a model with age-group interactions, along the lines
    of what you did in PS 1. Define the same four age groups, and
    estimate the same model, but interact the age-group dummies with
    TLyes rather than e. Estimate the model by OLS. Omit the control
    variables.*

```{r outcome var creation}
# adding outcome variables of interest to data set

# number of months receiving afdc created earlier (nmrecyr12)

ftp_merged %<>% 
  # number of quarters of employment
   mutate(nqemp12 = 
            # number of quarters y1 + number of quarter y2
            kempq2t5 + kempq6t9
  )
```

```{r age group creation}
# 4 age groups: less than 3, 3-8, 9-14 (36mo) or 9-15 (24mo), 15-19 (36mo) or 16-19 (24mo)

# alter ftp_merged
ftp_merged %<>% 
  # adding age group indicators
  mutate(
    # age group 1: less than 3
    ag1 = ifelse(yngchage <3, 1, 0),
    # age group 2: 3-8
    ag2 = ifelse(yngchage >= 3 & yngchage <9, 1, 0),
    # age group 3: 9-14 36mo, 9-15 24mo
    ag3 = ifelse(
      # children 9 or older and under 15 (9-14) 
      (yngchage >=9 & yngchage < 15 &
         # if time limit = 36 months (imputed for control)
         ((TLyes == 0 & bf_clk36 == 1) | (TLyes == 1 & l_clk36 == 1)))
      # OR children 9 or older and under 16 (9-15)
      |( yngchage >=9 & yngchage < 16) &
        # if time limit = 24 months (imputed for controls)
        ((TLyes == 0 & bf_clk24 == 1) | (TLyes == 1 & l_clk24 == 1)), 1, 0),
    # age group 4: 15-19 36mo, 16-19 24mo
    ag4 = ifelse(
      # children 15 or older and under 20 (15-19)
      (yngchage >=15 & yngchage < 20 &
         # if time limit = 36 months (imputed for control)
         ((TLyes == 0 & bf_clk36 == 1) | (TLyes == 1 & l_clk36 == 1))) 
      # OR children 16 or older and under 20 (16-19)
      |( yngchage >=16 & yngchage < 20) &
        # if time limit = 24 months (imputed for controls)
    ((TLyes == 0 & bf_clk24 == 1) | (TLyes == 1 & l_clk24 == 1)), 1, 0)) %>%
  # and filter out nas
  filter(!is.na(ag1 + ag2 + ag3 + ag4)) %>% 
   # and adding age group-treatment interactions
  mutate(
    tl1 = TLyes * ag1,
    tl2 = TLyes * ag2,
    tl3 = TLyes * ag3,
    tl4 = TLyes * ag4
  )

```

```{r age group model nmrecyr, results = 'asis'}
# models by outcome
lm_afdc <- lm(nmrecyr12 ~ TLyes + ag1 + ag2 + ag3 + tl1 + tl2 + tl3, data = ftp_merged)

lm_fs <- lm(nmfs12 ~ TLyes + ag1 + ag2 + ag3 + tl1 + tl2 + tl3, data = ftp_merged)

lm_emp <-lm(nqemp12 ~ TLyes + ag1 + ag2 + ag3 + tl1 + tl2 + tl3, data = ftp_merged)

stargazer(lm_afdc, lm_fs, lm_emp, type = "html", title = "OLS regression on TLyes by age group", column.labels = c("AFDC  ", "Food Stamps  ", "Employed"), covariate.labels = c("TLyes", "Group 1", "Group 2", "Group 3", "Trt x Group 1", "Trt x Group 2", "Trt x Group 3"))

```

(10) *Now estimate the same model, but use interactions between e and
     the age-group dummies as instruments for the interactions between
     TLyes and the age-group dummies.*

```{r age group instruments, results = 'asis'}
# models by outcome
iv_afdc <- ivreg(nmrecyr12 ~ TLyes + ag1 + ag2 + ag3 + tl1 + tl2 + tl3 | e + ag1 + ag2 + ag3 + (e*ag1) + (e*ag2) + (e*ag3), data = ftp_merged)

iv_fs <- ivreg(nmfs12 ~ TLyes + ag1 + ag2 + ag3 + tl1 + tl2 + tl3 | e + ag1 + ag2 + ag3 + (e*ag1) + (e*ag2) + (e*ag3), data = ftp_merged)

iv_emp <-ivreg(nqemp12 ~ TLyes + ag1 + ag2 + ag3 + tl1 + tl2 + tl3 | e + ag1 + ag2 + ag3 + (e*ag1) + (e*ag2) + (e*ag3), data = ftp_merged)

stargazer(lm_afdc, iv_afdc, lm_fs, iv_fs, lm_emp, iv_emp, type = "html", title = "IV v OLS on TLyes by age group", column.labels = c("AFDC", "AFDC", "Food Stamps  ", "Food Stamps  ", "Employed", "Employed"), covariate.labels = c("TLyes", "Group 1", "Group 2", "Group 3", "Trt x Group 1", "Trt x Group 2", "Trt x Group 3"))

```

*10a. Comment on how estimating the model with age-group interactions
helps to satisfy the exclusion restriction.*

The exclusion restriction requires that the instrument (age-group
interactions with e) only impact the outcome via the endogenous variable
(age-group interactions with TLyes). This instrument isolates the impact
of actual time limit exposure on each age group, not perceived time
limit exposure on each group.

Estimating the model with age-group interactions allows for a more
heterogeneous treatment effect. The model can thus better predict the
treatment effect by age group, and consequently more closely identify
the exogenous variation caused by the instrument. Isolating the
instruments impact allows the model to satisfy the exclusion
restriction, as we can be more sure that the instrument is not
endogenous.

(11) *Discuss your estimated effects of time limits and how they compare
     with the estimates from question (9). Can you diagnose why the two
     sets of coefficients differ?*

For AFDC receipt, the IV model predicts the same sign as the OLS model,
but the magnitude varies. For example, the IV model predicts a smaller
decrease in number of months receiving AFDC than OLS among treatment
interactions with age group 1. But age groups 2 and 3 see a greater
decrease in the number of months receiving AFDC in the IV model.

For FS receipt, the IV model follows a similar pattern to AFDC receipt.
The model of the treatment-age group interaction for group 1 shows a
very slight increase, while groups 2 and 3 show a very large decrease in
the IV model as compared to OLS.

For employment, the IV model shows a small positive increase among the
treatment-age group interactions for group 1 and 2, but a minor decrease
for group 3.

We can assume these two sets of coefficients differ because the
instrument is more accurately isolating the effect of the time limits,
while OLS is regressing on the endogenous variable TLyes.
